---
title: "**Analytic pipeline of DNA methylation data in R code**"
author: "**Rosaria Tornisiello**"
date: "6/14/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
## Step 1 
Loading of minfi, a R library that is used to perform the analysis and setting of the working directory. Loading of raw data with minfi and creation of an object called RGset storing the RGChannelSet object. 
```{r message=FALSE, warning=FALSE}
suppressMessages(library(minfi))
setwd("/home/rosaria/Desktop/DRD/")
baseDir <- ("./Report_DRD/Input_data/")
targets <- read.metharray.sheet(baseDir)
RGset <- read.metharray.exp(targets = targets)
RGset
```
## Step 2 
Creation of the dataframes Red and Green to store the red and green fluorescences respectively.
```{r message=FALSE, warning=FALSE}
Red <- data.frame(getRed(RGset))
Green <- data.frame(getGreen(RGset))
```
## Step 3 
Retrieving of red and green fluorescence for the address *59625465*.
```{r message=FALSE, warning=FALSE}
probe_red <- Red[rownames(Red)=="59625465",]
probe_green <- Green[rownames(Green)=="59625465",]

probe_red
```
```{r message=FALSE, warning=FALSE}
probe_green
```
|       Sample      | Red fluor | Green fluor | Type | 
|-------------------|-----------|-------------|------|
|X5775278051_R01C01 |  11585    |    935      |  II  |
|X5775278051_R04C02 |  14282    |    845      |  II  |
|X5775278078_R02C01 |  11253	  |    666	    |  II  |
|X5775278078_R05C01 |  11494    |    785      |  II  |
|X5775278078_R05C02 |  11254	  |    652	    |  II  |
|X5930514034_R01C02 |  11152    |    306	    |  II  |
|X5930514035_R04C02 |  11625	  |    584	    |  II  |
|X5930514035_R06C02 |  13019    |    668      |  II  |

### Optional
Checking in the manifest file if the address corresponds to a Type I or a Type II probe.
```{r message=FALSE, warning=FALSE}
load('./Illumina450Manifest_clean.RData')
Illumina450Manifest_clean[Illumina450Manifest_clean$AddressA_ID=="59625465",]
```
As it is possible to see in the output, the address corresponds to a Type II probe so no color channel is specified. Moreover, there is a correspondence only for the address A and not for the address B since Type II never have it.
## Step 4
Creation of the object MSet.raw
```{r message=FALSE, warning=FALSE}
MSet.raw <- preprocessRaw(RGset)
MSet.raw
```
## Step 5
Performing the following quality checks:

- QCplot
```{r message=FALSE, warning=FALSE}
qc <- getQC(MSet.raw)
plotQC(qc)
```
This Quality Control plot shows good results since all the point (samples) are localized above the diagonal.

- Checking the intensity of negative controls using minfi.
```{r message=FALSE, warning=FALSE}
controlStripPlot(RGset, controls="NEGATIVE")
```
The colors in this plot are inverted due to an error in the package. The written colors in the heading are correct. It is evident that all the negative controls have an intensity lower than 1000, so they are all good.

- Calculation of the detection pValues; for each sample, probes that have a detection p-value higher than the threshold 0.01 are computed and reported in the table below.
```{r message=FALSE, warning=FALSE}
detP <- detectionP(RGset)
failed <- detP>0.01
summary(failed)
```
|Sample           |Failed positions|
|5775278051_R01C01|323             |
|5775278051_R04C02|260             |
|5775278078_R02C01|312             |
|5775278078_R05C01|485             |
|5775278078_R05C02|465             |
|5930514034_R01C02|123             |
|5930514035_R04C02|60              |
|5930514035_R06C02|149             |

## Step 6
Calculation of raw beta and M values and plotting of the densities of mean methylation values. 
First the samples are divided in DS and WT.
```{r message=FALSE, warning=FALSE}
csv <- read.csv("./Report_DRD/Input_data/Samplesheet_report_2020.csv")
wt <- csv[csv$Group=="WT", "Basename"]
ds <- csv[csv$Group=="DS", "Basename"]
wtSet <- MSet.raw[,colnames(MSet.raw) %in% wt]
dsSet <- MSet.raw[,colnames(MSet.raw) %in% ds]
```
Beta and M are computed for both groups.
```{r message=FALSE, warning=FALSE}
wtBeta <- getBeta(wtSet)
wtM <- getM(wtSet)
dsBeta <- getBeta(dsSet)
dsM <- getM(dsSet)
```
Mean of beta and M are computed for both groups, discarding the NA values.
```{r message=FALSE, warning=FALSE}
mean_wtBeta <- apply(wtBeta,MARGIN=1,mean,na.rm=T)
mean_dsBeta <- apply(dsBeta,MARGIN=1,mean,na.rm=T)
mean_wtM <- apply(wtM,MARGIN=1,mean,na.rm=T)
mean_dsM <- apply(dsM,MARGIN=1,mean,na.rm=T)
```
The density distributions are computed for both beta and M.
```{r message=FALSE, warning=FALSE}
d_mean_wtBeta <- density(mean_wtBeta)
d_mean_dsBeta <- density(mean_dsBeta)
d_mean_wtM <- density(mean_wtM)
d_mean_dsM <- density(mean_dsM)
```
Finally the density plots are generated.
```{r message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
plot(d_mean_wtBeta,main="Density of Beta Values",col="orange")
lines(d_mean_dsBeta,main="Density of Beta Values",col="purple")
plot(d_mean_dsM,main="Density of M Values",col="orange")
lines(d_mean_dsM,main="Density of M Values",col="purple")
```
## Step 7
Normalization of the data using the function preprocessSWAN and comparison between raw data and normalized data. The output is a plot with 6 panels in which, for both raw and normalized data, there are the density plots of beta mean values according to the chemistry of the probes, the density plot of beta standard deviation values according to the chemistry of the probes and the boxplot of beta values.

First, the Illumina450Manifest_clean is divided in two dataframes, containing only type I (dfI) or type II (dfII) probes:
```{r message=FALSE, warning=FALSE}
dfI <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="I",]
dfI <- droplevels(dfI)
dfII <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="II",]
dfII <- droplevels(dfII)
```
In the beta matrix, which contains the beta values, the names of the probes are stored in the rownames of the matrix. The following code subsets the beta matrix in order to retain only the rows whose name is in the first column of dfI or in the first column of dfII.
```{r message=FALSE, warning=FALSE}
beta <- getBeta(MSet.raw)
beta_I <- beta[rownames(beta) %in% dfI$IlmnID,]
beta_II <- beta[rownames(beta) %in% dfII$IlmnID,]
```
For each probe in the mean_of_beta_I and mean_of_beta_II matrices, the mean of beta values across the 8 samples and  the density distribution of the 2 vectors of mean values are computed:

```{r message=FALSE, warning=FALSE}
mean_of_beta_I <- apply(beta_I,1,mean)
mean_of_beta_II <- apply(beta_II,1,mean)
d_mean_of_beta_I <- density(mean_of_beta_I,na.rm=T)
d_mean_of_beta_II <- density(mean_of_beta_II,na.rm=T)
```
The standard deviations of beta values are computed as well as theire densities:
```{r message=FALSE, warning=FALSE}
sd_of_beta_I <- apply(beta_I,1,sd,na.rm=T)
sd_of_beta_II <- apply(beta_II,1,sd,na.rm=T)
d_sd_of_beta_I <- density(sd_of_beta_I)
d_sd_of_beta_II <- density(sd_of_beta_II)
```
Data are normalized usign the preprocessSWAN function: Subset-quantile Within Array Normalisation (SWAN) for the Illumina Infinium HumanMethylation450 platform. It allows Infinium I and II type probes on a single array to be normalized together. 
```{r message=FALSE, warning=FALSE}
preprocessSWAN_results <- preprocessSWAN(RGset)
```
And the same process is repeated for the normalized data: calculation of beta, mean of beta, density of beta, standard deviation, mean of standard deviation and density of standard deviation
```{r message=FALSE, warning=FALSE}
beta_preprocessSWAN <- getBeta(preprocessSWAN_results)
beta_preprocessSWAN_I <- beta_preprocessSWAN[rownames(beta_preprocessSWAN) %in% dfI$IlmnID,]
beta_preprocessSWAN_II <- beta_preprocessSWAN[rownames(beta_preprocessSWAN) %in% dfII$IlmnID,]
mean_of_beta_preprocessSWAN_I <- apply(beta_preprocessSWAN_I,1,mean)
mean_of_beta_preprocessSWAN_II <- apply(beta_preprocessSWAN_II,1,mean)
d_mean_of_beta_preprocessSWAN_I <- density(mean_of_beta_preprocessSWAN_I,na.rm=T)
d_mean_of_beta_preprocessSWAN_II <- density(mean_of_beta_preprocessSWAN_II,na.rm=T)
sd_of_beta_preprocessSWAN_I <- apply(beta_preprocessSWAN_I,1,sd)
sd_of_beta_preprocessSWAN_II <- apply(beta_preprocessSWAN_II,1,sd)
d_sd_of_beta_preprocessSWAN_I <- density(sd_of_beta_preprocessSWAN_I,na.rm=T)
d_sd_of_beta_preprocessSWAN_II <- density(sd_of_beta_preprocessSWAN_II,na.rm=T)
```
Now it is possible to plot the results:
```{r message=FALSE, warning=FALSE}
par(mfrow=c(2,3))
plot(d_mean_of_beta_I,col="blue",main="raw beta")
lines(d_mean_of_beta_II,col="red")
plot(d_sd_of_beta_I,col="blue",main="raw sd")
lines(d_sd_of_beta_II,col="red")
boxplot(beta)
plot(d_mean_of_beta_preprocessSWAN_I,col="blue",main="preprocessSWAN beta")
lines(d_mean_of_beta_preprocessSWAN_II,col="red")
plot(d_sd_of_beta_preprocessSWAN_I,col="blue",main="preprocessSWAN sd")
lines(d_sd_of_beta_preprocessSWAN_II,col="red")
boxplot(beta_preprocessSWAN)
```
Comparing the density distributions of row beta and processSWAN beta means, it is evident that the type II probe peak is shifted toward the center and better aligned with the type I probe peak in the normalized data. Moreover, type I probes have higher density at low methylation levels and type II probes have higher density at high methylatilation levels. Regarding the standard deviation distributions, clearly, in the normalized data plot, the two peaks are more similar than in the raw data. Through the box plots it is easy to observe that the beta values fluctuate more across the 8 samples in the row data than in the normalized data.

## Step 8 
Performing a PCA analysis appling the function prcomp on the matrix of beta values obtained from normalized data in step 7.
```{r message=FALSE, warning=FALSE}
pca_results <- prcomp(t(beta_preprocessSWAN),scale=T)
summary(pca_results)
```
```{r message=FALSE, warning=FALSE}
plot(pca_results)
```
The PCA results plot reveals that the variance is almost equally distributed among the principal component going from 2 to 7; the firs PC covers the higher proportion of variance while the 8th one contains a variance equal to 0. 

```{r message=FALSE, warning=FALSE}
palette(c("orange","purple"))
plot(pca_results$x[,1], pca_results$x[,2],cex=2,pch=2,col=pheno$Group,xlab="PC1",ylab="PC2",xlim=c(-1000,1000),ylim=c(-1000,1000))
text(pca_results$x[,1], pca_results$x[,2],labels=rownames(pca_results$x),cex=0.5,pos=1)
legend("bottomright",legend=levels(pheno$Group),col=c(1:nlevels(pheno$Group)),pch=2)
```
## Step 9
Using the matrix of normalized beta values generated in step 7, this step identifies differentially methylated probes between group DS and group WT using the Mann-Whytney normalization (a non-parametric test).
```{r message=FALSE, warning=FALSE}
pheno <- read.csv("./Report_DRD/Input_data/Samplesheet_report_2020.csv",header=T, stringsAsFactors=T)
My_mannwhitney_function <- function(x) {
  wilcox <- wilcox.test(x~ pheno$Group)
  return(wilcox$p.value)
} 
pValues_wilcox <- apply(beta_preprocessSWAN, 1, My_mannwhitney_function)
```
Now it is possible to generate a dataframe with all the beta values and P-values
```{r message=FALSE, warning=FALSE}
final_wilcox <- data.frame(beta_preprocessSWAN, pValues_wilcox)
final_wilcox_sorted <- final_wilcox[order(final_wilcox$pValues_wilcox),]
summary(final_wilcox_sorted)
```
## Step 10 
Application of multiple test correction, setting a significant threshold of 0.05. The first step is the creation of a vector storing the p-values (starting from the matrix in which the rows are ordered according the p-value, from the smallest to the largest). Then the Bonferroni correction and the Benjamini & Hochberg corrections are applied. Finally, a dataframe and a boxplot are generated.
```{r message=FALSE, warning=FALSE}
raw_pValues <- final_wilcox_sorted[,9]
corrected_pValues_Bonf <- p.adjust(raw_pValues,"bonferroni")
corrected_pValues_BH <- p.adjust(raw_pValues,"BH")
final_wilcoxon_corrected <- data.frame(final_wilcox_sorted, corrected_pValues_BH, corrected_pValues_Bonf)
summary(final_wilcoxon_corrected)
```

```{r message=FALSE, warning=FALSE}
boxplot(final_wilcoxon_corrected[,9:11])
```
The following code identifies differentially methylated probes considering nominal pValues, BH correction, and Bonferroni correction.
```{r message=FALSE, warning=FALSE}
dim(final_wilcox_sorted[final_wilcox_sorted$pValues_wilcox<=0.05,])
dim(final_wilcox_sorted[final_wilcox_sorted$corrected_pValues_BH<=0.05,])
dim(final_wilcox_sorted[final_wilcox_sorted$corrected_pValues_Bonf<=0.05,])
```
The first number for each line of the output is the number of rows, so the number of differentially methylated probes considering the nominal p-value (22270 probes), the BH correction (0 probes) and the Bonferroni correction (0 probes).

## Step 11
Production of an heatmap of the top 100 differentially mehtylated probes. By default the function heatmap.2 uses the Euclidean distance as measure of distance and the complete linkage as linkage method for the hierarchical clustering. The first heatmap is generated using the default parameters.
```{r message=FALSE, warning=FALSE}
library("gplots")
input_heatmap=as.matrix(final_wilcox_corrected[1:100,1:8])
colorbar <- c("green","green","orange","orange","green","green","orange","orange")
heatmap.2(input_heatmap,col=terrain.colors(100),Rowv=T,Colv=T,dendrogram="both",key=T,ColSideColors=colorbar,density.info="none",trace="none",scale="none",symm=F)
```
